---
title: "ML prediction of Mortality"
output: html_document
date: "2026-01-31"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
## Loading the libraries
library(kableExtra)
library(bench)
library(dplyr)
library(knitr)
library(data.table)
#install.packages("arrow")
library(arrow)
library(tidyr)
```

```{r}
ldata_fread <- fread("tu_new.csv")
#str(ldata_fread)

###We used Apache Arrow and the Parquet format because the NLMS Tobacco Use dataset is large (~493,000 observations with many variables), and base R data handling becomes inefficient and memory-intensive at this scale.

write_parquet(ldata_fread, "tu_new.parquet")
ds <- open_dataset("tu_new.parquet")
```

```{r}

df <- ds %>%
  dplyr::select(curruse,race, age, sex,ms,histatus,adjinc,inddea,smokstat,smok100,agesmk,everuse) %>%
  collect()

ds_mod <- df %>%
  mutate(
    curruse = as.character(curruse),
    curruse_recode = case_when(
      grepl("1", curruse) ~ 1,  # any “1” anywhere → Yes
      curruse == "22222" ~ 0,   # all “2”s → No
      TRUE ~ NA_real_           # any “9” or invalid → Missing
    )
    )%>%
    mutate(
    everuse = as.character(everuse),
    everuse_recode = case_when(
      grepl("1", everuse) ~ 1,          # any “1” anywhere → Yes
      everuse == "22222" ~ 0,           # all “2”s → No
      TRUE ~ NA_real_                   # any “9” or invalid → Missing
    )
    )%>%
   mutate(
    ms_recode = case_when(
      ms == 1 ~ 1,                # Married
      ms %in% c(2, 3,4,5) ~ 0,     # Widowed, Separated,Divorced, Never married → Not married
      TRUE ~ NA_real_             # Missing or unknown
    )
  )


```

```{r}
set.seed(123)
### Fitting the ML prediction model
n <- nrow(ds_mod)
n
train_index <- sample(1:n, size = 0.7 * n)

train_data <- ds_mod[train_index, ]
nrow(train_data)
#head(train_data)
test_data  <- ds_mod[-train_index, ]
nrow(test_data)
#View(test_data)

#class(train_data$smokstat)
#unique(train_data$smokstat)
#unique(test_data$smokstat)

train_data$smokstat <- factor(train_data$smokstat)
#train_data$smokstat
test_data$smokstat <- factor(test_data$smokstat)
#test_data$smokstat
#sum(complete.cases(train_data))
#sum(complete.cases(test_data))

### Model1=============================

### training
## Exposure : Smokestat, Reference = Never smoker
model_smoking <- glm(inddea~ smokstat + age + factor(race) +sex+ factor(histatus) +adjinc + factor(ms_recode),
                  family = binomial,
                  data = train_data)

## Testing
test_data <- na.omit(test_data)

# Now predict
pred_prob_m1 <- predict(model_smoking, newdata = test_data, type = "response")
#pred_prob_m1 

library(pROC)
roc(test_data$inddea, pred_prob_m1)
plot(roc(test_data$inddea, pred_prob_m1), col="blue", main="ROC Curve Model 1")


## Model 2 ==================================================================
### Exposure : Duration (Smokestat + agesmk, Reference = everyday smoker)

model_smoking_duration <- glm(inddea ~ smokstat+ agesmk+ age + factor(race)+factor(sex)+factor(histatus)+adjinc+factor(ms_recode),
                  family = binomial,
                  data = train_data)

pred_prob_m2 <- predict(model_smoking_duration, newdata = test_data, type = "response")
#pred_prob_m2



roc1 <- roc(test_data$inddea, pred_prob_m1)
roc2 <- roc(test_data$inddea, pred_prob_m2)
roc.test(roc1, roc2)  # test if the difference in AUC is significant



```

```{r}
# Install if needed
#install.packages("caret")
library(caret)
set.seed(123)

## Cross Validation

ds_mod$inddea <- factor(ds_mod$inddea, levels = c(0,1))
n <- nrow(ds_mod)
n
train_index <- sample(1:n, size = 0.7 * n)

train_data <- ds_mod[train_index, ]
#nrow(train_data)
#head(train_data)
test_data  <- ds_mod[-train_index, ]
#nrow(test_data)

train_data$smokstat <- factor(train_data$smokstat)
#train_data$smokstat
test_data$smokstat <- factor(test_data$smokstat)
#test_data$smokstat

# Define 5-fold CV
train_control <- trainControl(
  method = "cv",   # cross-validation
  number = 5,      # 5 folds
  classProbs = TRUE, # needed if you want probabilities
  summaryFunction = twoClassSummary # allows AUC calculation
)

train_data <- na.omit(train_data)
test_data <- na.omit(test_data)


train_data$inddea <- factor(train_data$inddea, levels = c(0,1), labels = c("No", "Yes"))
test_data$inddea  <- factor(test_data$inddea, levels = c(0,1), labels = c("No", "Yes"))

# Train logistic regression with 5-fold CV
set.seed(123) # for reproducibility
cv_model1 <- train(
  inddea ~  smokstat + age + factor(race) +sex+ factor(histatus) +adjinc + factor(ms_recode),
  data = train_data,
  method = "glm",
  family = "binomial",
  trControl = train_control,
  metric = "ROC" # optimize based on AUC
)

# View CV results
cv_model1

cv_model2 <- train(
  inddea ~ smokstat+ agesmk+ age + factor(race)+factor(sex)+factor(histatus)+adjinc+factor(ms_recode),
  data = train_data,
  method = "glm",
  family = "binomial",
  trControl = train_control,
  metric = "ROC" # optimize based on AUC
)

# View CV results
cv_model2





```

| Model     | ROC        | Sensitivity | Specificity |
|-----------|------------|-------------|-------------|
| cv_model1 | 0.8461     | 0.9986      | 0.0307      |
| cv_model2 | **0.8465** | 0.9985      | **0.0345**  |

-   **ROC (AUC)** =\~ 0.85 → The models have **excellent discrimination**, meaning it can rank individuals from low to high mortality risk quite well.

-   **Sensitivity** \~ 0.999 → The models are **almost perfect at identifying deaths** (True Positive Rate)

-   **Specificity** = 0.03 → Very low, meaning it **misclassifies many survivors as deaths** (False Positive Rate is high).

    **Note:** This is expected if your dataset is **highly imbalanced**, e.g., deaths are a small fraction of total samples. Logistic regression tends to predict the **majority class poorly for specificity** unless you handle class imbalance.

```{r}
## predicting the probabilities
pred_probs1 <- predict(cv_model1, newdata = test_data, type = "prob")$Yes
pred_probs2 <- predict(cv_model2, newdata = test_data, type = "prob")$Yes

## Checking ROC
rocCV1 <- roc(test_data$inddea, pred_probs1)
rocCV2 <- roc(test_data$inddea, pred_probs2)
roc.test(rocCV1, rocCV2)

# ROC Plot
plot(rocCV1, col = "blue", main = "ROC Curve: Model Comparison")
lines(rocCV2, col = "red")
legend("bottomright", legend = c("Model 1: smokstat", "Model 2: smokstat+agesmk"),
       col = c("blue", "red"), lwd = 2)
```

The AUCs of SmokingStatus model`rocCV1` (0.852) and SmokingStatus + SmokingDuration model `rocCV2` (0.853) are very similar, and the small but statistically significant difference (p ≈ 3.44e-5) indicates `rocCV2` has a slightly higher discriminatory performance.

The models have **excellent discrimination**, meaning it can rank individuals from low to high mortality risk quite well

```{r}
# Convert outcome to numeric
obs_numeric <- ifelse(test_data$inddea == "Yes", 1, 0)


# 90th percentile cutoff (Top 10%)
cutoff_10_1 <- quantile(pred_probs1, 0.90)
cutoff_10_2 <- quantile(pred_probs2, 0.90)

top10_1 <- ifelse(pred_probs1 >= cutoff_10_1, 1, 0)
top10_2 <- ifelse(pred_probs2 >= cutoff_10_2, 1, 0)

# Baseline mortality rate
baseline_rate <- mean(obs_numeric)
baseline_rate 

# Event rate in top 10%
top10_rate_1 <- mean(obs_numeric[top10_1 == 1])
top10_rate_2 <- mean(obs_numeric[top10_2 == 1])
top10_rate_1
top10_rate_2 

lift_10_1 <- top10_rate_1 / baseline_rate
lift_10_2 <- top10_rate_2 / baseline_rate

lift_10_1
lift_10_2
```

```{r}
## Compare Top 5%, 10%, 20% Together

compute_lift <- function(pred, obs, percentile) {
  
  cutoff <- quantile(pred, 1 - percentile)
  top_group <- pred >= cutoff
  
  baseline <- mean(obs)
  top_rate <- mean(obs[top_group])
  
  lift <- top_rate / baseline
  
  return(c(
    percentile = percentile * 100,
    baseline_rate = baseline,
    top_rate = top_rate,
    lift = lift
  ))
}

lift5_1  <- compute_lift(pred_probs1, obs_numeric, 0.05)
lift10_1 <- compute_lift(pred_probs1, obs_numeric, 0.10)
lift20_1 <- compute_lift(pred_probs1, obs_numeric, 0.20)


lift5_2  <- compute_lift(pred_probs2, obs_numeric, 0.05)
lift10_2 <- compute_lift(pred_probs2, obs_numeric, 0.10)
lift20_2 <- compute_lift(pred_probs2, obs_numeric, 0.20)

rbind(lift5_1, lift10_1, lift20_1)
rbind(lift5_2, lift10_2, lift20_2)

percentiles1 <- seq(0.01, 0.5, by = 0.01)

lift_values <- sapply(percentiles1, function(p) {
  compute_lift(pred_probs1, obs_numeric, p)["lift"]
})

par(mfrow = c(1, 2))

p1 <- plot(percentiles1 * 100, lift_values, type = "l",
     xlab = "Top Risk Percentile (%)",
     ylab = "Lift",
     main = " SmokingStatus Model")

percentiles2 <- seq(0.01, 0.5, by = 0.01)

lift_values <- sapply(percentiles2, function(p) {
  compute_lift(pred_probs2, obs_numeric, p)["lift"]
})

p2 <- plot(percentiles2 * 100, lift_values, type = "l",
     xlab = "Top Risk Percentile (%)",
     ylab = "Lift",
     main = " SmokingStatus + Duration Model ")

par(mfrow = c(1, 1))
```

> Lift analysis demonstrated strong risk stratification performance. The top 10% highest predicted risk individuals exhibited approximately five-fold higher mortality compared to the overall population, with even greater concentration observed in the top 5% risk group.

```{r}

## Model1 Calibration 

cal_data1 <- data.frame(
  obs = test_data$inddea,
  pred = pred_probs1
)

cal_data2 <- data.frame(
  obs = test_data$inddea,
  pred = pred_probs2
)

## Method 1 : Manual Binning

cal_plot <- cal_data1 %>%
  mutate(bin = ntile(pred, 20)) %>%
  group_by(bin) %>%
  summarise(
    mean_pred = mean(pred),
    obs_rate = mean(as.numeric(obs) - 1),
    n = n()
  )

ggplot(cal_plot, aes(x = mean_pred, y = obs_rate)) +
  geom_point(size = 2) +
  geom_line() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(x = "Mean Predicted Probability",
       y = "Observed Mortality Rate") +
  ggtitle("Smoking Status Model") +
  theme_minimal() +  # optional: cleaner look
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
  )
       


## methood 2: Loess
library(ggplot2)

ggplot(cal_data1, aes(x = pred, y = as.numeric(obs) - 1)) +
  geom_smooth(method = "loess", se = FALSE) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(x = "Predicted Probability", y = "Observed Mortality")

```

```{r}
## Model2 Calibration 



cal_data2 <- data.frame(
  obs = test_data$inddea,
  pred = pred_probs2
)

## Method 1 : Manual Binning

cal_plot2 <- cal_data2 %>%
  mutate(bin = ntile(pred, 20)) %>%
  group_by(bin) %>%
  summarise(
    mean_pred = mean(pred),
    obs_rate = mean(as.numeric(obs) - 1),
    n = n()
  )

ggplot(cal_plot2, aes(x = mean_pred, y = obs_rate)) +
  geom_point(size = 2) +
  geom_line() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(x = "Mean Predicted Probability",
       y = "Observed Mortality Rate")+
  ggtitle("Smoking Status + Smoking Duration Model") +
  theme_minimal() +  # optional: cleaner look
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
  )

## methood 2: Loess
library(ggplot2)

ggplot(cal_data2, aes(x = pred, y = as.numeric(obs) - 1)) +
  geom_smooth(method = "loess", se = FALSE) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(x = "Predicted Probability", y = "Observed Mortality")
```

```{r}
# Convert Yes/No to numeric
obs_numeric <- ifelse(test_data$inddea == "Yes", 1, 0)

brier1 <- mean((pred_probs1 - obs_numeric)^2)
brier2 <- mean((pred_probs2 - obs_numeric)^2)

brier1
brier2

```

Model calibration was evaluated using grouped (decile-based) calibration analysis. Predicted probabilities were divided into equal-frequency bins, and observed event rates were calculated within each bin. Manual binning was chosen due to the large sample size and imbalanced outcome distribution, which can lead to instability in nonparametric smoothing approaches (e.g., LOESS), particularly in extreme probability ranges. Grouped calibration provides stable estimation of observed risk, improves interpretability, and aligns with recommended practices for reporting clinical prediction models. Both models demonstrated good overall calibration, with low Brier scores (Model 1: 0.0447; Model 2: 0.0446), indicating accurate probability estimation. The inclusion of smoking duration yielded only marginal improvement in overall prediction error.
